Spontaneous breaking of multipole symmetries

### 2022-03-08 multipoleSSB/manuscript.tex
starting referee response

\documentclass[pra,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4-2}

@@

@@

Reference~\cite{Gromov2019} includes discussion of some multipole groups; we will review a few here. The simplest case is the maximal multipole group $\mathcal{M}^a_\text{max}$, which includes all shifts by polynomials of degree $a$ or less. Individual polynomials can be written as
\begin{align}
P_c^{\gamma} = \mu^{\gamma}_{I_c}x^{I_c}, \label{eqn:basis}
\end{align}
where $\gamma$ is an index that runs over polynomials of degree $c$ and $c\le a$. The composite index $I_c = \{i_1,\dots,i_c\}$  expands as $x^{I_c}=x^{i_1}\cdots x^{i_c}$, and similarly for $\partial_{I_c}$. Each  matrix  $\mu^{\gamma}$ is fully symmetric on its $I_c$ indices. This group also includes all translations and rotations.

An example of a multipole group that contains all translations and rotations but is \emph{not} the maximal multipole group is the group generated by shifts of the form
\begin{align}
\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{\beta} P^{\beta}_2,
\end{align}
where the degree-0 polynomial is $P_0^0=1$. The other polynomials are
\begin{align}
P_1^{i} = x^i,\quad \quad P_2^{\beta} &= \mu^{\beta}_{ij} x^i x^j,
\end{align}
where $\mu^{\beta}$ are a basis for the rank-2 traceless symmetric $d\times d$ matrices. Let us call this group $\mathcal{M}^2_{\text{sym}}$. Recall that the maximal quadrupole group $\mathcal{M}^2_{\text{max}}$ is already only built from symmetric matrices $\mu$. The tracelessness condition thus only removes one polynomial from the set.

@@

For $a>0$, $K_a$ can be split into multiple terms while still remaining rotationally invariant. For example, the kinetic term for $\mathcal{M}^1_\max$ is 
\begin{align}
K_1[\phi(x)] = g_1 (\partial^2\phi)^2 + g_2 \sum_{ij} (\partial_i \partial_j \phi)^2.
\end{align}
For the special cases of $g_2=0$ or $g_2=-g_1$, the symmetry group expands to $\mathcal{M}^2_\text{sym}$ or $\mathcal{M}^2_\text{tr}$, respectively. However, for generic $g_1,g_2$ the symmetry group is $\mathcal{M}^1_\max$. Similar statements can be made for larger $a$.

@@

We are now ready to describe the generalized Mermin-Wagner argument in the case where an arbitrary maximal multipole group is spontaneously broken down to the trivial subgroup. This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow. Here, we will discuss both thermal and $T=0$ systems.

Consider first a clean system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
Then the kinetic term is proportional to 
\begin{align}
K[\phi(x)] = (\partial_{I_{a+1}} \phi)^2. \label{eqn:adisp}
\end{align} 
Now consider nucleating domains of linear size $L$ in which the expectation value of the order parameter --- namely $\langle e^{i\phi}\rangle$ --- varies by an amount of order 1. Since the order parameter is continuous there will be a thick domain ``wall" over which the order parameter changes gradually from its value in the background phase to its value in the domain. The thickness will be of order $L$.
Equation~\ref{eqn:adisp} tells us the  energy cost of the field changing its value will be minimized if it varies as a polynomial with degree $a+1$. In that case the energy density will scale as $K_k \sim k^{2(a+1)} \sim L^{-2(a+1)}$. Integrating that energy density over a region of size $L$ gives a total energy cost to nucleating domains of $L^{d-2(a+1)}$.

@@

We can also reproduce this argument by considering correlation functions. As a warm-up, consider the standard case where a monopole U(1) symmetry is spontaneously broken, leading to a Goldstone boson $\phi(x)$. The correlation function
\begin{align}
C(x) = \left\langle \phi(x) \phi(0) \right\rangle = \int \frac{d^dk}{(2\pi)^d} \frac{e^{i k \cdot x}}{k^2}, 
\end{align}
diverges when $d\le 2$. Correlation functions should not diverge, so the interpretation is that the Goldstone boson fluctuates strongly enough that the field $\phi(x)$ cannot be well defined. In turn, this tells us the symmetry could not have been broken. This is the classical Mermin-Wagner argument. We should note that, since we only care about long-distance divergences, we don't need to include the complex exponential as long as we only look for divergence at small $k$. We will thus drop this dependence in future calculations.

@@

Consider a situation in which a dipolar symmetry $\mcm^1_{\rm max}$ is spontaneously broken to its monopolar subgroup $\mcm^0$, and specialize to the case of $d=3$ and $T>0$. In the spirit of Goldstone's theorem, the most naive thing to do when analyzing the symmetry broken phase would be to write down an action involving a set of $d=3$ Goldstone fields $\theta_j$, with $\t_j$ transforming linearly under the dipolar part of the symmetry group, but remaining invariant under the unbroken monopole subgroup, so that $\theta_j \mapsto \theta_j + \beta_j$ under a transformation parameterized by $P(x) = \alpha + \beta_j x^j$. In this case, the IR field theory for the putative symmetry-breaking phase would be of the form 
\begin{align}
\label{vector} \mathcal{L}= \sum_j  (\partial_\tau \theta_j)^2 + \sum_{i,j} g_{ij} (\partial_i \theta_j)^2.
\end{align}
As we are working in $d=3$ and at $T>0$, the theory \eqref{vector} indeed has SSB for the dipole symmetry, with $e^{i\t_j}$ developing long-range order.

@@

A general formal framework for understanding the counting of Goldstone modes in the present context can be developed by following the analysis of \cite{watanabe2013redundancies}, but here we will simply content ourselves with a heuristic physical argument. For simplicity, we will focus on systems in which the total symmetry group $\mcm_{\rm max}^a$ is spontaneously broken down to some nontrivial subgroup.

Let us define the compressibilities $\kappa_c$ by 
\be \label{kappadef} \kappa_c^{I_c} = \frac{d\langle n_c^{I_c}\rangle}{d\mu_c^{I_c}},\ee 
where $n_c$ is the number density of $c$-pole charges, $\mu_c$ is a conjugate chemical potential, and $I_c = \{i_1,\dots,i_c\}$ is a composite index as before.

Let $b\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e. consider a situation in which $\kappa_c>0$ for all $b\leq c\leq a$, while $\kappa_c=0$ for all $c<b$. Then the minimal IR Lagrangian for the symmetry broken phase looks schematically like 
\be \label{generall} \mathcal{L} = \kappa_b (\partial_\tau \phi_{I_b})^2 + g_{J_{a-b+1}}(\partial_{J_{a-b+1}}\phi_{I_b})^2,\ee 
where 
$\p_{J_{a-b+1}} \equiv \p_{j_1} \cdots \p_{j_{a-b+1}}$, and where $\phi_{I_b}$ is a field that shifts by a polynomial of degree $a-b$ under the action of $\mmax a$. The compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_b}$ is the momentum conjugate to $\phi_{I_b}$ in the quantum theory, and hence correlation functions of fluctuations in $n_b$ are determined by those of $\p_\tau \phi_{I_b}$.

A necessary condition for $\mmax b$ to be spontaneously broken is that $\kappa_b \neq 0$. Indeed, if $\mmax b$ is spontaneously broken then $n_b$ creates Goldstones when acting on the ground state, thereby ensuring that the density-density correlator appearing in the calculation of \eqref{kappadef} must be nonzero at long wavelengths.

However, a nonzero $\kappa_b$ does not {\it necessarily} imply that $\mmax b$ is spontaneously broken. $\kappa_b\neq 0$ means that the system possesses gapless modes created by $n_b$, but it is possible for fluctuations to be strong enough so that these modes cannot be associated with the Goldstone modes of a spontaneously broken $\mmax b$. Thus the theory \eqref{generall} does not necessarily describe a phase where $\mmax b$ is spontaneously broken, and the exact pattern of symmetry breaking depends on the spatial dimension and whether or not $T$ is nonzero. For example, we can again consider the theory in \eqref{singlescalar} at $d=3, T>0$: even though the charge compressibility $\kappa_0$ (the coefficient of $(\p_\tau\phi)^2$ in \eqref{singlescalar}) is nonzero, $\phi$ fluctuates strongly enough that the monopole symmetry is not spontaneously broken.

@@

In the previous subsection, we saw that the number of Goldstone modes in the symmetry-breaking theory is determined by the smallest degree of multipole charge $b$ such that $\kappa_b\neq0$. We now turn to determining the symmetry breaking pattern that occurs for a given choice of $b$, spatial dimension $d,$ and temperature $T$, thereby allowing us to formulate a generalized version of the Mermin-Wagner theorem for the case of partial multipole breaking.

Since $\kappa_c=0$ if $c<b$, none of the subgroups $\mmax {c<b}$ can be spontaneously broken. Suppose then that the symmetry is spontaneously broken down to $\mmax {c}$, where $c\geq b$. To find the allowed values of $c$ consistent with this assumption, we need to calculate correlation functions of operators of the form $\exp(i\p_{J_{c-b}}\phi_{I_b})$, these being the operators transforming nontrivially under $\mmax c$ with the longest-range correlation functions. We have 
\be\ba \langle \p_{J_{c-b}}\phi_{I_b}(x) \p_{J_{c-b}}\phi_{I_b}(0)\rangle & \sim \int \frac{d\omega \, dk\, k^{d-1+2(c-b)}}{\omega^2 + k^{2(a-b+1)}} \\ 
& \sim \int dk\, k^{d-1+2(c-b) - \varsigma_T(a-b+1)}, \ea\ee 
where we have defined the function $\varsigma_T$ such that $\varsigma_T = 1$ if $T=0$ and $\varsigma_T = 2$ if $T>0$. Consequently, we find that $\mmax c$ can be spontaneously broken only in dimensions $d$ such that 
\be \label{pbreak_mw} d>2(b-c) + \varsigma_T(a-b+1).\ee 
For a given choice of symmetry breaking pattern (captured by $a$ and $c$) and a given choice of which compressibilities are nonzero (captured by $b$), the quantity on the RHS determines the lower critical dimension for symmetry breaking, thereby generalizing the Mermin-Wagner theorem to the partial breaking case. 
As a sanity check on \eqref{pbreak_mw}, note that when we set $c=b=0$ (full multipole breaking) we recover the lower critical dimension of $\varsigma_T(a+1)$ obtained in the previous section. Also note that as expected, the lower critical dimensions in the quantum $(\varsigma_T = 1)$ and classical $(\varsigma_T = 2)$ cases are related by $d_{\rm cl} = d_{\rm q} + z$, where $z=a-b+1$ is the dynamical exponent of the Goldstone theory \eqref{generall}.

Finally, as an interesting example, consider the case where the dipolar group $(a=1)$ is broken down to the monopole group $(b=0,c=1)$. At $T=0$, we see from \eqref{pbreak_mw} that this is possible provided that $d>0$. Thus the partial breaking of dipole symmetry provides us with an example where a continuous symmetry can be spontaneously broken in $d=1$ at zero temperature.

@@

We can then compute the correlation function for $\phi(x)$,
\begin{align}
\langle \phi(0) \phi(x) \rangle &\sim \int d^dk \, K_k^{-2} \langle h_{-k} h_{k} \rangle e^{ik\cdot x},
\end{align}
where $\langle h_{-k} h_{k} \rangle e^{ik\cdot x}$ does not affect the divergence at small $k$, assuming the disorder is short range correlated. We can compare to Eqn.~\ref{eqn:correl} to see that in the presence of disorder that couples linearly to an order parameter fully breaking multipole symmetry, the critical dimension for a disordered system is twice the critical dimension for having direct full multipole symmetry breaking in a clean classical system.

@@

We now present a transparent quantum lattice example to illustrate the spontaneous breaking of multipolar symmetry. To this end, we take seriously the idea that ``dipole charges are monopole charges of a dipole symmetry," and include explicit secondary degrees of freedom that couple to dipole moments of the original degrees of freedom. This allows us to access phases corresponding to partial or complete symmetry breaking.

@@

The minimal Hamiltonian obeying these symmetries is  
\begin{align}
H_0  &= H_b + H_d + H_\text{int}\nn
H_b &= \sum_j \left[ \frac{U_b}{2}  n_j (n_j-1) - \mu_b \, n_j \right] \nn
&\qquad - \sum_{j,\mu,\nu} t_b \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n^d_{j,j+\mu} (n^d_{j,j+\mu} - 1) -\mu_d \, n^d_{j, j + \mu} \right]\nn
&\qquad  - \sum_{j,\mu,\nu} t_d \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
H_\text{int} &= \sum_{j,\mu} g\left[ b_j d_{j,j+\mu} b^\dag_{j+\mu} + H.C. \right],
\end{align}
where sums are taken over sites $j$ and lattice directions $\mu$ and $\nu$.  
For simplicity, let us set 
\begin{align}
\frac{\mu_b}{ U_b} = \frac{\mu_d }{ U_d} = \frac{1}{2},
\end{align}
so there is a robust insulating phase at weak hopping $t_b, t_d$~\cite{Fisheretal}. A detailed discussion of the phase diagram and phenomenology of this model will be given in~\cite{dbhm}; in the following we will simply discuss the symmetry breaking patterns realized in various different limits.

>>>>

\documentclass[prb,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4-2}

@@

The differences from the ordinary Mermin-Wagner theorem all come from the soft dispersions of the relevant Goldstone modes. Although soft Goldstone modes exist in other models, they generically only emerge at fine-tuned points. As we will see, the multipole symmetries make the soft modes natural, in the sense that the theories do not have to be fine-tuned. Rather, the soft modes must exist in the presence of various multipolar symmetries.

@@

Reference~\cite{Gromov2019} includes discussion of some multipole groups; we will review a few here. The simplest case is the maximal multipole group $\mathcal{M}^a_\text{max}$, which includes all shifts by polynomials of degree $a$ or less. Individual polynomials can be written as
\begin{align}
P_c^{\gamma} = \mu^{\gamma}_{I_c}x^{I_c}, \label{eqn:basis}
\end{align}
where $\gamma$ is an index that runs over polynomials of degree $c$ and $c\le a$. The composite index $I_c = \{i_1,\dots,i_c\}$  expands as $x^{I_c}=x^{i_1}\cdots x^{i_c}$, and similarly for $\partial_{I_c}$. Each  matrix  $\mu^{\gamma}$ is fully symmetric on its $I_c$ indices. This group also includes all translations and rotations. We will use $\mmax{-1}$ to denote the rotations and translations without any internal symmetry.

An example of a multipole group that contains all translations and rotations but is \emph{not} the maximal multipole group is the group generated by shifts of the form
\begin{align}
\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{\beta} P^{\beta}_2,
\end{align}
where the degree-0 polynomial is $P_0^0=1$. The other polynomials are
\begin{align}
P_1^{i} = x^i,\quad \quad P_2^{\beta} &= \mu^{\beta}_{ij} x^i x^j,
\end{align}
where $\mu^{\beta}$ are a basis for the traceless symmetric $d\times d$ matrices. Let us call this group $\mathcal{M}^2_{\text{sym}}$. Recall that the maximal quadrupole group $\mathcal{M}^2_{\text{max}}$ is already only built from symmetric matrices $\mu$. The tracelessness condition thus only removes one polynomial from the set.

@@

For $a>0$, $K_a$ can be split into multiple terms while still remaining rotationally invariant. For example, the kinetic term for $\mathcal{M}^1_\max$ is 
\begin{align}
K_1[\phi(x)] = g_1 \big(\sum_{i} \partial_i \partial_i \phi\big)^2 + g_2 \sum_{i \ne j} (\partial_i \partial_j \phi)^2.
\end{align}
For the special cases of $g_2=0$ or $g_2=0$, the symmetry group expands to $\mathcal{M}^2_\text{sym}$ or $\mathcal{M}^2_\text{tr}$, respectively. However, for generic $g_1,g_2$ the symmetry group is $\mathcal{M}^1_\max$. Similar statements can be made for larger $a$.

Although we have been discussing field theories in the continuum, we will often care about theories that emerge from the low-energy degrees of freedom of a lattice model. In ordinary theories with monopole and lattice rotation symmetries, the lowest-order derivative term that can emerge from the lattice theory is $(\partial_i \phi)^2$, which is symmetric under the larger group of continuous rotations. Any terms that are not invariant under the continuous rotation group are irrelevant, so the continuous symmetry emerges. For multipole symmetries, leading-order terms may contain lattice anisotropies, so that the continuous rotation group does not emerge. For this reason we will only require that $\mmax{a}$ contain discrete rotation symmetries.

@@

We are now ready to describe the generalized Mermin-Wagner argument in the case where an arbitrary maximal multipole group is spontaneously broken down to the trivial subgroup. This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow. Here, we will discuss both thermal and $T=0$ systems. We will start with a heuristic argument in terms of domain wall energies before discussing a more careful diagnosis of spontaneous symmetry breaking in terms of correlation functions.

Consider first a clean system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
Then the kinetic term is proportional to 
\begin{align}
K[\phi(x)] = (\partial_{I_{a+1}} \phi)^2. \label{eqn:adisp}
\end{align} 
Symmetry breaking field configurations will be of the form $\phi(x) = c + c_ix^i + \dots+ c_{I_a}x^{I_a}$, so there is no kinetic energy.

Now consider nucleating domains of linear size $L'$ in which the field takes on a new configuration $\phi(x) = c' + c'_ix^i + \dots + c'_{I_a}x^{I_a}$. In the thick domain ``wall" surrounding the domain, the coefficients will vary from their background unprimed values to the primed values in the domain. The equations of motion from~\eqref{eqn:adisp} say that $\phi$ should vary as a degree-$(2a+1)$ polynomial in this region.

Let us align the domain wall in the $x^1$ direction with thickness $L$, so that $\phi$ takes its unprimed values at $x^1=0$ and its primed values at $x^1=L$. We can build a degree-$(2a+1)$ polynomial by including factors of $(x^1)^{2a+1}/L^{2a+1}$. Terms with fewer powers of $L$ in the denominator will not be seen by the kinetic energy. For large $L$, terms with more powers of $L$ in the denominator will be subleading. Thus the energy density in the domain wall scales as $K\sim L^{-2(a+1)}$. Integrating over a spherical shell gives a total energy cost of
\begin{align}
E &\sim \left( (L'+L)^d -(L')^d \right) L^{-2(a+1)} \nonumber\\
&\sim L^{d-2(a+1)},
\end{align}
where the last line comes from setting $L\sim L'$, which minimizes the energy cost.

@@

We can reproduce this argument by considering correlation functions. As a warm-up, consider the standard case where a monopole U(1) symmetry is spontaneously broken, leading to a Goldstone boson $\phi(x)$. The correlation function
\begin{align}
C(x) = \left\langle \phi(x) \phi(0) \right\rangle = \int \frac{d^dk}{(2\pi)^d} \frac{e^{i k \cdot x}}{k^2}, 
\end{align}
diverges when $d\le 2$. Correlation functions should not diverge, so the interpretation is that the Goldstone boson fluctuates strongly enough that the field $\phi(x)$ cannot be well defined. In turn, this tells us the symmetry could not have been broken. This is the classical Mermin-Wagner argument. We should note that, since we only care about long-distance divergences, we don't need to include the complex exponential as long as we only look for divergence at small $k$. We will thus drop this dependence in future calculations.

@@

Consider a situation in which a dipolar symmetry $\mcm^1_{\rm max}$ is spontaneously broken to its monopolar subgroup $\mcm^0_{\rm max}$, and specialize to the case of $d=3$ and $T>0$. In the spirit of Goldstone's theorem, the most naive thing to do when analyzing the symmetry broken phase would be to write down an action involving a set of $d=3$ Goldstone fields $\theta_j$, with $\t_j$ transforming linearly under the dipolar part of the symmetry group, but remaining invariant under the unbroken monopole subgroup, so that $\theta_j \mapsto \theta_j + \beta_j$ under a transformation parameterized by $P(x) = \alpha + \beta_j x^j$. In this case, the IR field theory for the putative symmetry-breaking phase would be of the form 
\begin{align}
\label{vector} \mathcal{L}= \sum_j  (\partial_\tau \theta_j)^2 + \sum_{i,j} g_{ij} (\partial_i \theta_j)^2.
\end{align}
As we are working in $d=3$ and at $T>0$, the theory \eqref{vector} indeed has SSB for the dipole symmetry, with $e^{i\t_j}$ developing long-range order.

Of course, such a Goldstone field would break continuous rotation symmetries, but recall that we are only requiring that each maximal multipole group contain lattice rotation symmetries. We can then say that the theory~\eqref{vector} spontaneously breaks $\mmax{1}$ to $\mmax{0}$, in $d=3$ and at $T>0$.

@@

A general formal framework for understanding the counting of Goldstone modes in the present context can be developed by following the analysis of \cite{watanabe2013redundancies}, but here we will simply content ourselves with a heuristic physical argument. For simplicity, we will focus on systems in which the total symmetry group $\mcm_{\rm max}^a$ is spontaneously broken down to some nontrivial subgroup $\mcm_{\rm max}^b$. As mentioned previously, $\mmax{b}$ might only contain discrete rotation symmetries, so we will only concern ourselves with the spontaneous breaking of the polynomial generators. We will indicate that a generator corresponding to a polynomial of degree $c$ is spontaneously broken by saying that $\mmax{c}$ is broken.

Determining the counting of the Goldstone modes requires introducing the compressibilities $\kappa_c$, defined by 
\be \label{kappadef} \kappa_c^{I_c} = \frac{d\langle n_c^{I_c}\rangle}{d\mu_c^{I_c}},\ee 
where $n_c$ is the number density of $c$-pole charges, $\mu_c$ is a conjugate chemical potential, and $I_c = \{i_1,\dots,i_c\}$ is a composite index as before.

A necessary condition for $\mmax c$ to be spontaneously broken is that $\kappa_c \neq 0$. Indeed, if $\mmax c$ is spontaneously broken then $n_c$ creates Goldstones when acting on the ground state, thereby ensuring that the density-density correlator appearing in the calculation of \eqref{kappadef} must be nonzero at long wavelengths. Since we are assuming that $\mmax c$ is spontaneously broken for all $c>b$, we must accordingly have $\kappa_{c>b} \neq 0$.

However, $\kappa_c \neq 0$ does not {\it necessarily} imply that $\mmax c$ is spontaneously broken: $\kappa_c\neq 0$ means that the system possesses gapless modes created by $n_c$, but it is possible for fluctuations to be strong enough so that these modes cannot be associated with the Goldstone modes of a spontaneously broken $\mmax c$ (this is of course what happens in the QLRO phase of the 1+1D XY model). Therefore, we are allowed to have $\kappa^c \neq 0$ even if $c \leq b$, without contradicting our assumption about the symmetry breaking pattern.

Let then $s\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e. consider a situation in which $\kappa_c>0$ for all $s\leq c\leq a$, while $\kappa_c=0$ for all $c<s$. 
Given $s$, the minimal IR Lagrangian looks schematically like 
\be \label{generall} \mathcal{L} = \kappa_s (\partial_\tau \phi_{I_s})^2 + g_{J_{a+1}}(\partial_{J_{a-s+1}}\phi_{I_s})^2,\ee where 
$\p_{J_{a-s+1}} \equiv \p_{j_1} \cdots \p_{j_{a-s+1}}$, and where $\phi_{I_s}$ is a field that shifts by a polynomial of degree $a-s$ under the action of $\mmax a$ (the compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_s}$ is the momentum conjugate to $\phi_{I_s}$ in the quantum theory, and hence correlation functions of fluctuations in $n_s$ are determined by those of $\p_\tau \phi_{I_s}$).  If $s=b+1$, then $\phi_{I_s}$ may develop LRO while leaving $\mmax{b}$ invariant.

Note that the number of modes in the IR theory is controlled by $s$, and {\it not} by $b$. In order for $s< b+1$ to be consistent with our assumed pattern of symmetry breaking, it must be the case that the fluctuations of the fields in \eqref{generall} are strong enough so as to prevent $\mmax {s}$ from spontanoeusly breaking. The novelty provided by multipolar symmetries is that they offer an easy way of allowing for $s$ to be smaller than $b+1$, and thus provide an easy way of realizing multiple IR theories that have the same symmetry breaking pattern, but contain different numbers of modes. This is illustrated by the examples of the theories \eqref{vector} ($s=1$) and \eqref{singlescalar} ($s=0$), which have the same symmetry breaking pattern ($b=0$) in $d=3$ and at $T>0$.

@@

In the previous subsection, we saw that the number of Goldstone modes in the symmetry-breaking theory is determined by the smallest degree of multipole charge $s$ such that $\kappa_s\neq0$. We now turn to determining the symmetry breaking pattern that occurs for a given choice of $s$, spatial dimension $d,$ and temperature $T$, thereby allowing us to formulate a generalized version of the Mermin-Wagner theorem for the case of partial multipole breaking.

Since $\kappa_c=0$ if $c<s$, none of the subgroups $\mmax {c<s}$ can be spontaneously broken. Suppose then that the subgroup $\mmax{c}$ is spontaneously broken, where $c\geq s$. To find the allowed values of $c$ consistent with this assumption, we need to calculate correlation functions of operators of the form $\exp(i\p_{J_{c-s}}\phi_{I_s})$, these being the operators transforming nontrivially under $\mmax c$ with the longest-range correlation functions. We have 
\be\ba \langle \p_{J_{c-s}}\phi_{I_s}(x) \p_{J_{c-s}}\phi_{I_s}(0)\rangle & \sim \int \frac{d\omega \, dk\, k^{d-1+2(c-s)}}{\omega^2 + k^{2(a-s+1)}} \\ 
& \sim \int dk\, k^{d-1+2(c-s) - \varsigma_T(a-s+1)}, \ea\ee 
where we have defined the function $\varsigma_T$ such that $\varsigma_T = 1$ if $T=0$ and $\varsigma_T = 2$ if $T>0$. Consequently, we find that $\mmax c$ can be spontaneously broken only in dimensions $d$ such that 
\be d>2(s-c) + \varsigma_T(a-s+1).\ee 
The preserved subgroup is then the largest multipole group $\mmax{b}$ that is not spontaneously broken.

For a given choice of symmetry breaking pattern (captured by $a$ and $b$) and a given choice of which compressibilities are nonzero (captured by $s$), the critical dimension for symmetry breaking is
\begin{align}
d_\text{lc} = 2(s - b-1) + \varsigma_T (a-s+1) \label{pbreak_mw} 
\end{align}
generalizing the Mermin-Wagner theorem to the partial breaking case. 
As expected, the lower critical dimensions in the quantum $(\varsigma_T = 1)$ and classical $(\varsigma_T = 2)$ cases are related by $d_{\rm cl} = d_{\rm q} + z$, where $z=a-s+1$ is the dynamical exponent of the Goldstone theory \eqref{generall}.

If we only care about the pattern of symmetry breaking, $\mmax{a}\rightarrow \mmax{b}$, the critical dimension can be found by setting $s=0$, so that
\begin{align}
d_\text{lc} = \varsigma_T (a+1) -2 (b+1). \label{pbreak_mw_nos}
\end{align}
As a sanity check on \eqref{pbreak_mw_nos}, note that for full multipole breaking, $b=-1$ (recall that $\mmax{-1}$ contains only translations and rotations), we recover the lower critical dimension of $\varsigma_T(a+1)$ obtained in the previous section. 
In any higher dimension, such a symmetry breaking pattern will be possible. However, $s$ must always be chosen so as to give the correct preserved subgroup. As emphasized earlier, the number of Goldstone modes depends on $s$, not $b$.

As an interesting example, consider the case where the dipolar group $(a=1)$ is broken down to the monopole group $(b=0)$. At $T=0$, we see from \eqref{pbreak_mw_nos} that this is possible provided that $d>0$. Thus the partial breaking of dipole symmetry provides us with an example where a continuous symmetry can be spontaneously broken in $d=1$ at zero temperature.

@@

We can then compute the correlation function for $\phi(x)$,
\begin{align}
\langle \phi(0) \phi(x) \rangle &\sim \int d^dk \, K_k^{-2} \langle h_{-k} h_{k} \rangle e^{ik\cdot x},
\end{align}
where $\langle h_{-k} h_{k} \rangle e^{ik\cdot x}$ does not affect the divergence at small $k$, assuming the disorder is short range correlated. We can compare to~\eqref{eqn:correl} to see that in the presence of disorder that couples linearly to an order parameter fully breaking multipole symmetry, the critical dimension for a disordered system is twice the critical dimension for having direct full multipole symmetry breaking in a clean classical system.

@@

We now present a transparent quantum lattice example to illustrate the spontaneous breaking of multipolar symmetry. We will not derive further results from this model. Rather, we apply our previous results to this model to understand how various phases are available in different dimensions. Reference~\cite{dbhm} gives detailed descriptions of the various phases and their transitions.

@@

The minimal Hamiltonian obeying these symmetries is  
\begin{align}
H_0  &= H_b + H_d + H_\text{int}\nn
H_b &= \sum_j \left[ \frac{U_b}{2}  n_j (n_j-1) - \mu_b \, n_j \right] \nn
&\qquad - \sum_{j,\mu,\nu} t_b \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n^d_{j,j+\mu} (n^d_{j,j+\mu} - 1) -\mu_d \, n^d_{j, j + \mu} \right]\nn
&\qquad  - \sum_{j,\mu,\nu} t_d \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
H_\text{int} &= \sum_{j,\mu} g\left[ b_j d_{j,j+\mu} b^\dag_{j+\mu} + H.C. \right],
\end{align}
where sums are taken over sites $j$ and lattice directions $\mu$ and $\nu$.  
For simplicity, let us set 
\begin{align}
\frac{\mu_b}{ U_b} = \frac{\mu_d }{ U_d} = \frac{1}{2},
\end{align}
so there is a robust insulating phase at weak hopping $t_b, t_d$~\cite{Fisheretal}. A detailed discussion of the phase diagram and phenomenology of this model is given in~\cite{dbhm}; in the following we will simply discuss the symmetry breaking patterns realized in various different limits.
